Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-08-15 analysis/paper.tex
Finalise resubmission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Open Quaternary}

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in the
terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 82 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.

The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. The models performance in
predicting the occurrence of specific species at archaeological sites is
highly variable, but in aggregate the predictions are coherent and align
with broad-scale trends in the ubiquity and relative abundance of
species in the archaeological record.
\end{abstract}

@@

In this paper we present a complementary, deductive approach based on
ecological niche modelling. Rather than inferring environmental
conditions from preserved physical evidence, we predict the ranges of
individual species relevant to human subsistence based on a model of
their current environmental niche and simulations of past palaeoclimate.
Though hypothetical, this gives us an independent line of evidence on
past ecologies that is independent of the environmental archaeological
and palaeoclimatic records. This means that the ancient data can be
reserved for assessing the model's ability to `hindcast' past
conditions. In this sense, discrepencies between the two records are
perhaps the most interesting result, as they indicate processes
affecting one or both records that are not fully accounted for and
therefore generate new questions. Our computational approach is also
readily scaled up, allowing us to model spatially-explicit
palaeodistributions for a large number of species, for the whole region,
under multiple past climatologies.

@@

Ecological niche modelling or species distribution modelling is widely
used by ecologists to predict the geographic range of a species based on
a set of environmental predictors \citep{FranklinMiller2009}.
Essentially, it involves combining records of where an organism has been
observed with environmental data (climate, topography, etc.) for those
locations to model the range of environmental values at which that
species -- its environmental niche. This model can then be used to
predict the range of the organism in question either in the same or a
different environment. \citet{TownsendPetersonSoberon2012} suggests
reserving the term `species distribution modelling' for when the method
is used to recover the verifiable range of a species in a real and
existing environment, and using `ecological niche modelling' as the
broader term covering hypothetical or predictive applications -- a
convention we follow here when referring to predictive or `hindcast'
models of past ranges. Within this overarching framework, ecological
niche modelling encompasses a wide range of applications and a variety
of potential environmental predictors, modelling approaches, and
methodologies, which we will not attempt to review here.

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited
\citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
has been used most extensively to model the range of humans and other
hominin species
\citep[e.g.][]{BenitoEtAl2017, YousefiEtAl2020, BanksEtAl2021, YaworskyEtAl2024a, YaworskyEtAl2024b, GuranEtAl2024},
especially in the Palaeolithic. This overlaps with what archaeologists
usually call generically `predictive modelling'
\citep{VerhagenWhitley2020}---or more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguished from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

@@

\caption{\label{fig-region}Late Epipalaeolithic \& Pre-Pottery Neolithic
archaeological sites used to generate modelled flora}

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Based on current understandings, we
assume that plant-based subsistence during this period was broad,
geographically- and temporally-varied, and reflects a gradual,
geographically decentred, and nonlinear tranistion to greater reliance
on cultivars (i.e.~agriculture, see Section~\ref{sec-bg}). Our starting
point was a list of 66 taxa (Table~\ref{tbl-results-summary}) comprising
the identified species observed at at least 3 Late
Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous
study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. This was based on dataset collated from
three previously published regional archaeobotanical databases: ADEMNES
\citep{ADEMNES}, ORIGINS \citep{ORIGINS}, and COMPAG
\citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
on][]{ColledgeEtAl2004}; \citealp{ShennanConolly2007}}. We did not
attempt to distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

@@

\begin{verbatim}
Rows: 1 Columns: 11
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (5): key, doi, license, status, downloadLink
dbl  (3): size, totalRecords, numberDatasets
dttm (3): created, modified, eraseAfter

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

Georeferenced occurrence data for West Eurasia between 0 and 60° of
latitude was obtained from the Global Biodiversity Information Facility
(GBIF) using via and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. The GBIF dataset
\citep{GBIFDerivedData} excluded fossil occurrences, recorded absences,
and records with missing or dubious coordinates. Although niche models
have reasonable predictive power even with small training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we did
not attempt to model 3 taxa with less than 40 usable occurrences,
following recommendations for niche models generally and Random
Forest-based models specifically
\citep{StockwellPeterson2002, LuanEtAl2020}. Multiple records of the
same taxon at the same coordinate were discarded because they do not
impart information to the model. The resulting cleaned dataset used to
train our niche models comprises \ensuremath{3.387196\times 10^{6}} from
4764 constituent datasets.

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia used to train models (N=1409939}

@@

GBIF is currently the best available general-purpose occurrence dataset
for the West Asia region, its coverage is uneven both geographically and
from species to species. Figure~\ref{fig-occ-map} shows clearly that the
Southern Levant, and Israel specifically, is significantly more densely
sampled than other parts of West Asia.

@@

\begin{verbatim}
bio_1, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
\end{verbatim}

\begin{verbatim}
Warning in parse_netcdf_meta(meta_data, get_names(x)): NAs introduced by
coercion
\end{verbatim}

\begin{verbatim}
bio_10, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_11, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_12, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_13, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_14, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_15, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_16, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_17, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_18, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_19, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_2, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_3, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_4, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_5, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_6, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_7, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_8, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

\begin{verbatim}
bio_9, 
\end{verbatim}

\begin{verbatim}
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
Vertical dimension.
Warning in CPL_read_gdal(as.character(x), as.character(options),
as.character(driver), : NAs introduced by coercion
\end{verbatim}

@@

\begin{verbatim}
Warning: There were 57 warnings in `mutate()`.
The first warning was:
i In argument: `occ = map(occ, bind_extract, st_as_stars(hack(climate,
  "period", "cur")))`.
Caused by warning in `CPL_get_metadata()`:
! GDAL Message 1: dimension #0 (period) is not a Time or Vertical dimension.
i Run `dplyr::last_dplyr_warnings()` to see the 56 remaining warnings.
\end{verbatim}

@@

We trained Random Forest models for 63 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in Section~\ref{sec-predictors}.
Substituting the ``current'' climate predictors for those derived from
palaeoclimatic simulations \citep{BrownEtAl2018}, we could then generate
hindcast predictions for reconstructed past environments in 4 key
climate periods -- a total of 252 modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the
appendix and accompanying material.

@@

\renewcommand\refname{References}
  \bibliography{references.bib}

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint]{elsarticle}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Open Quaternary}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in the
terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{jar@hum.ku.dk} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 65 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential niches of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
Our models performed well against independent contemporary test data,
but their ability to predict the occurrence of specific species at
archaeological sites was much more variable, probably reflecting a
tendency of the method to underestimate the species' fundamental niche.
Nevertheless, the majority of species are predicted to have had more
restricted geographic distributions under past climate conditions
compared to today. Crop progenitors and several wild food species
modelled are modelled to have been concentrated in the Levant and, to a
lesser extent, Cyprus and Western Anatolia, suggesting these regions may
have served as glacial \emph{refugia}. The average size of species'
niche shrunk by an average of c.~25\% from the terminal Pleistocene to
the Early Holocene, indicating that economically significant plants were
adapted to cryo-arid conditions and did not, as often assumed, initially
respond positively to the `ameliorated' climate of the Holocene.
\end{abstract}

@@

In this paper we present a complementary, deductive approach based on
ecological niche modelling. Rather than inferring environmental
conditions from preserved physical evidence, we predict the potential
niche of individual species relevant to human subsistence based on a
model of their current environmental niche and simulations of past
palaeoclimate. Though hypothetical, this gives us an independent line of
evidence on past ecologies that is independent of the environmental
archaeological and palaeoclimatic records. This means that the ancient
data can be reserved for assessing the model's ability to `hindcast'
past conditions. In this sense, discrepencies between the two records
are perhaps the most interesting result, as they indicate processes
affecting one or both records that are not fully accounted for and
therefore generate new questions. Our computational approach is also
readily scaled up, allowing us to model spatially-explicit
palaeodistributions for a large number of species, for the whole region,
under multiple reconstructed climate scenarios.

@@

Ecological niche modelling or species distribution modelling is widely
used by ecologists to mathematically predict the geographic distribution
of a species from a sample of occurrences
\citep{FranklinMiller2009, SilleroEtAl2021}. Essentially, it involves
combining records of where an organism has been observed with
environmental data (climate, topography, etc.) for those locations to
model the range of environmental values at which that species -- its
environmental niche. This model can then be used to predict the
potential niche of the organism in question either in the same or a
different environment. \citet{TownsendPetersonSoberon2012} suggests
reserving the term `species distribution modelling' for when the method
is used to recover the verifiable range of a species in a real and
existing environment, and using `ecological niche modelling' as the
broader term covering hypothetical or predictive applications -- a
convention we follow here when referring to predictive or `hindcast'
models of past niches. Within this overarching framework, ecological
niche modelling encompasses a wide range of applications and a variety
of potential environmental predictors, modelling approaches, and
methodologies, which we will not attempt to review here.

Ecological niche modelling has long been of interest to archaeologists
as both a means of exploring the biological niche of humans and for
reconstructing the past environments they inhabited
\citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
has been used most extensively to model the niche of humans and other
hominin species
\citep[e.g.][]{BenitoEtAl2017, YousefiEtAl2020, BanksEtAl2021, YaworskyEtAl2024a, YaworskyEtAl2024b, GuranEtAl2024},
especially in the Palaeolithic. This overlaps with what archaeologists
usually call generically `predictive modelling'
\citep{VerhagenWhitley2020}---or more precisely `site distribution
modelling'---which is essentially the same approach as (and often
borrows methodologies from) ecological niche modelling but applied to
the occurrence of archaeological sites. Here what is modelled is not
strictly a biological niche alone, but also aspects of human geography,
taphonomy, and archaeological visibility. These applications can be
distinguished from `palaeoecological niche modelling', where the object
of model remains, as in ecology, a non-human biological niche.

@@

\caption{\label{fig-region}Map of the study region (West Asia, grey box)
with locations of Late Epipalaeolithic and Pre-Pottery Neolithic
archaeobotanical assemblages.}

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Based on current understandings, we
assume that plant-based subsistence during this period was broad,
geographically- and temporally-varied, and reflects a gradual,
geographically decentred, and nonlinear tranistion to greater reliance
on cultivars (i.e.~agriculture, see Section~\ref{sec-bg}). Our starting
point was a list of 68 taxa (Table~\ref{tbl-results-summary}) comprising
the identified species observed at at least 3 Late
Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous
study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. This was based on dataset collated from
three previously published regional archaeobotanical databases: ADEMNES
\citep{ADEMNES}, ORIGINS \citep{ORIGINS}, and COMPAG
\citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
on][]{ColledgeEtAl2004}; \citealp{ShennanConolly2007}}. We did not
attempt to distinguish between the source of the remains
\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
to a variety of preservational and recovery biases, so by no means were
all the species on our list consumed or even deliberately collected by
people. However, we assume that there presence at a site of human
settlement at least implies that they were part of the wider ecosystem
that supported habitation there.

@@

Georeferenced occurrence data for West Eurasia between 0 and 60° of
latitude was obtained from the Global Biodiversity Information Facility
(GBIF) using via and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. The GBIF dataset
\citep{GBIFDataDownload} excluded fossil occurrences, recorded absences,
and records with missing or dubious coordinates. Although niche models
have reasonable predictive power even with small training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we did
not attempt to model 3 taxa with less than 40 usable occurrences,
following recommendations for niche models generally and Random
Forest-based models specifically
\citep{StockwellPeterson2002, LuanEtAl2020}. Multiple records of the
same taxon at the same coordinate were discarded because they do not
impart information to the model. The resulting cleaned dataset used to
train our niche models comprises 3392920 occurrences from 4769
constituent datasets.

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Eurasia used to train models (N=1412083). Inset, right: prediction
region (West Asia).}

@@

GBIF is currently the best available general-purpose occurrence dataset
for the West Asia region, its coverage is uneven both geographically and
from species to species. The Southern Levant, and Israel specifically,
is significantly more densely sampled than other parts of West Asia
(Figure~\ref{fig-occ-map}).

@@

@@

@@

We trained Random Forest models for 65 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in Section~\ref{sec-predictors}.
Substituting the ``current'' climate predictors for those derived from
palaeoclimatic simulations \citep{BrownEtAl2018}, we could then generate
hindcast predictions for reconstructed past environments in 4 key
climate periods -- a total of 260 modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the
appendix and accompanying material.

@@

\renewcommand\refname{References}
\bibliography{references.bib,gbif.bib}
